---
title: "occupancy-modeling-basics"
author: "jt-miller"
date: "2023-11-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Occupancy Modeling Basics:
Based on James Paterson's Occupancy Model introduction lessons: https://jamesepaterson.weebly.com/blog/introduction-to-occupancy-models-in-r

#### What are Occupancy Models
Occupancy Models estimate where species occur (a biological process) while accounting for imperfect detection (an observation process). The reason for this is that there will always be a probability of detection associated with a myraid of factors such as observation variation, cryptic species, conditions, etc... <br> 
Without accounting for imperfect detection, our estimates of species presence and absence (P/A) for sites can be biased. <br>

Basic Occupancy Models estimate two parameters 
$$
p_i,_t \\
\Psi
$$
P sub i and t represents the probability of finding a species at time t if it is at site i 
Psi represents the probability that a species occurs at site i 

#### Building Basic Occupancy Models with the Unmarked Package
We'll be first looking at a single-season model with the assumption that the occupancy status (P/A) does not change during the study. This means that a species is _always_ either present or absent at each site, even when it is not detected. Dynamic occupancy models will touch on allowing occupancy status at a site to change through colonization and extinction. 

First we need data that describes on which surveys a species was detected. Using the Sample data that James Paterson simulated, lets use the unmarked package which has a function for organizing the detection data & covariates. 
```{r}
library(unmarked)

# Simulated detection history (100 sites with 10 visits each)
detection_history <- read.csv("/home/jt-miller/Soltis-lab/useful-code-depository/occupancy-modeling/data/raw/detection_history.csv",
                              row.names = "X" # First variable is actually a row labeler not data
                              )

# Take a look at the data 
head(detection_history)
```
Set up the data
```{r}
# Use the fxn unmarkedFrameOccu to make the data compatible for occupancy detections

sample.unmarkedFrame_simple <- unmarkedFrameOccu(
  y = as.matrix(detection_history) # y is a matrix with observed detection history
  # 0s and 1s one row per site, one column per survey
)

summary(sample.unmarkedFrame_simple) # give a nice little summary about our simulated data
```
Recall the two parameters of interest: _Detection_ & _Occupancy_. <br>
Each of these parameters is estimated using a linear model on the logit scale, which bounds the real estimate between 0 and 1. We need to bound the estimates becuase the probabilities must fall between 0 and 1. <br> 

When building our occupancy model, we are going to specify that we will start with an *intercept-only model* where there is only one estimate for detection probability and one estimate for occupancy probability. An intercept-only model in this case will say that the probability of a species being present is the same at each site, and the probability of detecting a species is the same during each survey
```{r}
# Build the basic single-season occupancy model with intercepts only (one estimate for detection, one for occupancy)
occu.m1 <- occu(formula = ~1 ~1, # detection covariate first, occupancy covariate second
                data = sample.unmarkedFrame_simple)

summary(occu.m1) # Shows the AIC, estimates on logit scale the SE and z-scores
```
Accounting for confidence intervals (95% CI) to get a real estimate of occupancy
```{r}
predict(occu.m1, 
        newdata = data.frame(site =1),
        type = "state")
```
And same thing for a real estimate of detection
```{r}
predict(occu.m1, 
        newdata = data.frame(site =1),
        type = "det")
```
Find equivalent to reverse logit
```{r}
boot::inv.logit(coef(occu.m1)[1]) # real estimate of occupancy
boot::inv.logit(coef(occu.m1)[2]) # real estimate of detection
```
Here, we've built a simple single season occupancy model and extracted the real parameters predicting the occupancy probability (~0.62) and the detection probability (~0.48). <br> 
This means that we predict the species will occur in 62% of the sites, and that our ability to detect a species when present will be about 48% of the time. <br>

This simplistic model lacks reality however; as we know that a species will be more likely to inhabit particular sites based on the underlying characteristics of that site. To accomplish this, we can add covariates to the model. 

#### Adding Covariates for Occupancy and Detection
There are two type of covariates: 
1. Observation-level covariates, these will differ for every combination of site & survey (such as the search effort per visit per site for example). Each one of these variables should be included as an element of a list, and each element of the list should have a row for every site and a column for every survey. For this example we will have 100 rows and 10 columns

2. Site-level covariates, these do not change through time, but will differ by site. E.g. would be measures of habitat type or other landscape metrics. Site covariates should be included in a df with a row for every site and column for each site variable. 
