---
title: "Intro-to-fxns"
author: "jt-miller"
date: "2023-12-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Art of R Programming: 1.3 Introduction to Functions
A function is a group of instructions that takes inputs and uses them to compute other values. 

E.g. Create a function that counts the odd numbers in a vector of integers
```{r}
x <- c(1,2,4) # create a vector of numbers

# Count the number of odd intergers in x 

oddcount <- function(x) { # first tell r we want to define the function 'oddcount' with one argument x, the { } denote the beginning and end of the fxn. 
  k <- 0 # initialize our count at zero
  for(n in x){  # set n to x[1], x[2], ... x[n]
    # where %% is the modulo operator (basically gives you the remainder of a division operation e.g. 31%%5 = 1) 
    if(n %% 2 == 1) # if the remainder is == 1 after being divided by 2
      k <- k + 1 #  then it must be odd therefore add one to k 
  }
  return(k)
}

oddcount(x) # one odd number in the vector x 
oddcount(c(1,4,5,1,5)) # 4 odd numbers in this example vector...

```
One thing of particular note is that R programming tends to avoid loops if possible, and when used to keep them as simple as possible 
```{r}
oddcount <- function(x) { # first tell r we want to define the function 'oddcount' with one argument x, the { } denote the beginning and end of the fxn. 
  k <- 0 # initialize our count at zero
  for(i in 1:length(x)){  # the good-old c++ way. 
    # where %% is the modulo operator (basically gives you the remainder of a division operation e.g. 31%%5 = 1) 
    if(x[i] %% 2 == 1) # if the remainder is == 1 after being divided by 2
      k <- k + 1 #  then it must be odd therefore add one to k 
  }
  return(k) # return the value k 
}

oddcount(x) # one odd number in the vector x 
oddcount(c(1,4,5,1,5)) # 4 odd numbers in this example vector...

```
This does create the same output, unless x has a length = 0, but is generally discouraged since its simpler and cleaner 

More terminology: 
```{r}
oddcount <- function(x) { # first tell r we want to define the function 'oddcount' with one argument x, the { } denote the beginning and end of the fxn. 
  k <- 0 # initialize our count at zero
  for(n in x){  # set n to x[1], x[2], ... x[n]
    # where %% is the modulo operator (basically gives you the remainder of a division operation e.g. 31%%5 = 1) 
    if(n %% 2 == 1) # if the remainder is == 1 after being divided by 2
      k <- k + 1 #  then it must be odd therefore add one to k 
  }
  return(k)
}

# of note, x is known as the formal argument (or formal parameter) of the function oddcount().
# in the following function call
oddcount(c(1,3,5)) # c(1,3,5) is known as the actual argument. These terms are used to differentiate the fact that x is really just a placeholder, while what actually goes into the function (c(1,3,5) in this case) is what is actually used in the computation.  

```

### Variable Scope

A variable that is only visable within the function body is local to that function. in oddcount(), k & n are the local variables. The key thing to know about local variables is that they disappear after the function returns, hence you cannot call 'n' 
```{r}
oddcount(c(1,2,4))
n # wont run
```
Variables created outside of functions are global variables, and are available everywhere. 
```{r}
f <- function(x) return(x+y) # note that we just are saying return whatever we input into the formal argument, x, + y
y <- 3 # here we assign y as a global variable 
f(5) # here we put the input of the x argument 
# we could also write a global variable within a function by using the superassignment operator '<<-' 
f <- function(x){
  y <<- 3 # globally assign y as 3 
return(x+y)
}
f(5) # when we call this function, note that y is populated into our global environment as 3 
```
### Default Arguments
```{r}
g <- function(x, y=2, z=T){...} # here we build a dummy fxn that has y and z inialized if the user does note provide input, 
g(12,z=FALSE) # x is a required actual argument, while y can be safely ignored and left to its default value of 2. Finally we can Overwrite the z's default status of T (boolean TRUE) with FALSE if we wish

# Side note: T and F can be used for TRUE and FALSE, but honestly I dont really care for this as it limits our variable naming style while introducing room for confusion. 
```


