---
title: "vectors"
author: "jt-miller"
date: "2024-06-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 2 Vectors

### Adding and Deleting Vector Elements
Vectors are stored like arrays in C, contiguously, therefore we cannot insert/delete elements. The size of a vecotr is determined at creation, you must reassign the vecotr in order to add or delete elements. 
```{r}
# how to add an element in the middle of a 4 element vector
x <- c(88,5,12,13)
x <- c(x[1:3], 168, x[4]) # insert 168 before 13, reassign vector
x
```
Note that here what actually happened is that we createed a new vector and storaged that vector within x. this has subtle requirement has implications on restricting speed and preformance in R (talked about later).

### Obtaining Length of a Vector 
Using the length() fxn
```{r}
x <- c(1,2,4)
length(x)
```
Suppose that we needed this for a more complex scenario: 
We want to have a fxn that determines the index of the first 1 value in the function's vector argument
Here's a non-efficient way of doing this:
```{r}
first1 <- function(x){
  for(i in 1:length(x)){
    if (x[i] == 1) break # break out of loop
  }
  return(i)
} 
# Note that this loop fails edge case that if the vector has a length of 0 it will fail. 
```

### Matrices and Arrays as Vectors

Arrays and matrices (and even lists) are just vectors with some extra class attributes. For example matrices are given the # of rows and columns. 
```{r}
m <- matrix(c(1,3,2,4), nrow = 2, ncol = 2)
m + 10:13 # note that R remebers that we're using matrix in variable m, therefore did matrix addition. 
```
### Declarations 
Unlike compiler languages, we do not need to declare variables (warn the interpreter/compiler) of a variabes existence prior to assigning it. Scripting languages typically do not require this meaning that the following is valid 
```{r}
z <- 3
```
When referencing specific

